---
title: "CodeBook for MSDS6306 - Case Study 2"
author: "Lavonnia Newman, Jeff Washburn, & Joseph Caguioa"
date: "2/25/2019"
output: 
  html_document:
documentclass: article
header-includes:
  - \newcommand{\fullline}{\noindent\makebox[\linewidth]{\rule{\textwidth}{0.4pt}}}
  - \renewcommand\familydefault{\sfdefault}
  - \newcommand{\bminione}{\begin{minipage}{0.75 \textwidth}}
  - \newcommand{\bminitwo}{\begin{minipage}{0.25 \textwidth}}
  - \newcommand{\emini}{\end{minipage}}
---
```{r , echo=FALSE, include=FALSE, warning=FALSE, message=FALSE, error=FALSE}
library("ggplot2")
library("pander")
```
```{r 'visualFunctions', echo=FALSE, include=FALSE, warning=FALSE, message=FALSE, error=FALSE}
ggAggHist <- getFromNamespace("ggAggHist", "dataMaid")
ggAggBarplot <- getFromNamespace("ggAggBarplot", "dataMaid")
```
# Data report overview
The dataset examined has the following dimensions:


---------------------------------
Feature                    Result
------------------------ --------
Number of observations       1470

Number of variables            35
---------------------------------




# **Codebook Summary Table**

## employee 

-------------------------------------------------------------------------------------   
Original Unedited            Variable             Class         # unique    Missing     
 Names                                           values                        
--------------------      --------------------  --------------  ----------- --------- 
Age                        **[Age]**             numeric             43       0.00 %                

Attrition                  **[Attrition]**      character            2        0.00 %                
    
BusinessTravel             **[BusTravel]**      character            3        0.00 %                

DailyRate                  **[DailyRate]**      numeric            886        0.00 %                

Department                 **[Department]**     character            3        0.00 %                

DistranceFromHOme          **[DistFromHme]**    numeric             29        0.00 %                

Education                  **[Education]**      numeric              5        0.00 %                

EducationField             **[EduField]**       character            6        0.00 %                

EmployeeCount              **[EmpCount]**       numeric              1        0.00 %                

EmployeeNumber             **[EmpId]**          numeric           1470        0.00 %                

EnvironmentSatisfaction    **[EnvironSat]**     numeric              4        0.00 %                

Gender                     **[Gender]**         character            2        0.00 %                

HourlyRate                 **[HourlyRate]**     numeric             71        0.00 %                

JobInvolvement             **[JobInvolvemt]**   numeric              4        0.00 %                

JobLevel                   **[JobLevel]**       numeric              5        0.00 %                

JobRole                    **[JobRole]**        character            9        0.00 %                

JObSatisfaction            **[JobSat]**         numeric              4        0.00 %                

MaritalStatus              **[MaritalStat]**    character            3        0.00 %                

MonnthlyIncome             **[MnthInc]**        numeric           1349        0.00 %                

MonthlyRate                **[MnthRate]**       numeric           1427        0.00 %                

NumCompaniesWorked         **[NumCoWorked]**    numeric             10        0.00 %                

Over18                     **[Over18]**         character            1        0.00 %                

Overtime                   **[Overtime]**       character            2        0.00 %                

PercentSalaryHike          **[PcntSalInc]**     numeric             15        0.00 %                

PerformanceRating          **[PerfRating]**     numeric              2        0.00 %                

RelationshipSatisfaction   **[RelationSat]**    numeric              4        0.00 %                

StandardHours              **[StandHours]**     numeric              1        0.00 %                

StockOptionLevel           **[StckOptLev]**     numeric              4        0.00 %                

TotalWorkingYears          **[TotalWrkYrs]**    numeric             40        0.00 %                

TrainingTimesLastYear      **[TrainPrevYr]**    numeric              7        0.00 %                

WorkLifeBalance            **[WorkLifeBal]**    numeric              4        0.00 %                

YearsAtCompany             **[YrsAtComp]**      numeric             37        0.00 %                

YearsInCurrentRole         **[YrsCurrRole]**    numeric             19        0.00 %                

YearsSinceLastPromotion    **[YrsSncPromo]**    numeric             16        0.00 %                

YearsWithCurrManager       **[YrsWtCurMgr]**    numeric             18        0.00 %                
-------------------------------------------------------------------------------------




## **Variable list**
### Age

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values          43

Median                           36

1st and 3rd quartiles        30; 43

Min. and max.                18; 60
-----------------------------------


\emini
\bminitwo
```{r 'Var-1-Age', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(15L, 20L, 25L, 30L, 
35L, 40L, 45L, 50L, 55L), xmax = c(20L, 25L, 30L, 35L, 40L, 45L, 
50L, 55L, 60L), ymin = c(0, 0, 0, 0, 0, 0, 0, 0, 0), ymax = c(28L, 
95L, 263L, 343L, 276L, 192L, 130L, 96L, 47L)), class = "data.frame", row.names = c(NA, 
-9L)), vnam = "Age")
```
\emini




\fullline

### Attrition

\bminione

-------------------------------------
Feature                        Result
------------------------- -----------
Variable type               character

Number of missing obs.        0 (0 %)

Number of unique values             2

Mode                             "No"
-------------------------------------


\emini
\bminitwo
```{r 'Var-2-Attrition', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggBarplot(data = structure(list(x = structure(1:2, .Label = c("No", 
"Yes"), class = "factor"), y = c(1233L, 237L)), class = "data.frame", row.names = c(NA, 
-2L)), vnam = "Attrition")
```
\emini


- Observed factor levels: \"No\", \"Yes\". 



\fullline

### BusTravel

\bminione

-------------------------------------------
Feature                              Result
------------------------- -----------------
Variable type                     character

Number of missing obs.              0 (0 %)

Number of unique values                   3

Mode                        "Travel_Rarely"
-------------------------------------------


\emini
\bminitwo
```{r 'Var-3-BusTravel', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggBarplot(data = structure(list(x = structure(1:3, .Label = c("Non-Travel", 
"Travel_Frequently", "Travel_Rarely"), class = "factor"), y = c(150L, 
277L, 1043L)), class = "data.frame", row.names = c(NA, -3L)), 
    vnam = "BusTravel")
```
\emini


- Observed factor levels: \"Non-Travel\", \"Travel_Frequently\", \"Travel_Rarely\". 



\fullline

### DailyRate

\bminione

-------------------------------------
Feature                        Result
------------------------- -----------
Variable type                 numeric

Number of missing obs.        0 (0 %)

Number of unique values           886

Median                            802

1st and 3rd quartiles       465; 1157

Min. and max.               102; 1499
-------------------------------------


\emini
\bminitwo
```{r 'Var-4-DailyRate', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(100L, 200L, 300L, 400L, 
500L, 600L, 700L, 800L, 900L, 1000L, 1100L, 1200L, 1300L, 1400L
), xmax = c(200L, 300L, 400L, 500L, 600L, 700L, 800L, 900L, 1000L, 
1100L, 1200L, 1300L, 1400L, 1500L), ymin = c(0, 0, 0, 0, 0, 0, 
0, 0, 0, 0, 0, 0, 0, 0), ymax = c(106L, 91L, 103L, 105L, 123L, 
119L, 86L, 107L, 107L, 90L, 111L, 105L, 123L, 94L)), class = "data.frame", row.names = c(NA, 
-14L)), vnam = "DailyRate")
```
\emini




\fullline

### Department

\bminione

----------------------------------------------------
Feature                                       Result
------------------------- --------------------------
Variable type                              character

Number of missing obs.                       0 (0 %)

Number of unique values                            3

Mode                        "Research & Development"
----------------------------------------------------


\emini
\bminitwo
```{r 'Var-5-Department', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggBarplot(data = structure(list(x = structure(1:3, .Label = c("Human Resources", 
"Research & Development", "Sales"), class = "factor"), y = c(63L, 
961L, 446L)), class = "data.frame", row.names = c(NA, -3L)), 
    vnam = "Department")
```
\emini


- Observed factor levels: \"Human Resources\", \"Research & Development\", \"Sales\". 



\fullline

### DistFromHme

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values          29

Median                            7

1st and 3rd quartiles         2; 14

Min. and max.                 1; 29
-----------------------------------


\emini
\bminitwo
```{r 'Var-6-DistFromHme', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(0, 2, 4, 6, 8, 10, 12, 
14, 16, 18, 20, 22, 24, 26, 28), xmax = c(2, 4, 6, 8, 10, 12, 
14, 16, 18, 20, 22, 24, 26, 28, 30), ymin = c(0, 0, 0, 0, 0, 
0, 0, 0, 0, 0, 0, 0, 0, 0, 0), ymax = c(419L, 148L, 124L, 164L, 
171L, 49L, 40L, 58L, 46L, 47L, 37L, 55L, 50L, 35L, 27L)), class = "data.frame", row.names = c(NA, 
-15L)), vnam = "DistFromHme")
```
\emini




\fullline

### Education

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values           5

Median                            3

1st and 3rd quartiles          2; 4

Min. and max.                  1; 5
-----------------------------------


\emini
\bminitwo
```{r 'Var-7-Education', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(1, 1.5, 2, 2.5, 3, 3.5, 
4, 4.5), xmax = c(1.5, 2, 2.5, 3, 3.5, 4, 4.5, 5), ymin = c(0, 
0, 0, 0, 0, 0, 0, 0), ymax = c(170L, 282L, 0L, 572L, 0L, 398L, 
0L, 48L)), class = "data.frame", row.names = c(NA, -8L)), vnam = "Education")
```
\emini




\fullline

### EduField

\bminione

-------------------------------------------
Feature                              Result
------------------------- -----------------
Variable type                     character

Number of missing obs.              0 (0 %)

Number of unique values                   6

Mode                        "Life Sciences"
-------------------------------------------


\emini
\bminitwo
```{r 'Var-8-EduField', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggBarplot(data = structure(list(x = structure(1:6, .Label = c("Human Resources", 
"Life Sciences", "Marketing", "Medical", "Other", "Technical Degree"
), class = "factor"), y = c(27L, 606L, 159L, 464L, 82L, 132L)), class = "data.frame", row.names = c(NA, 
-6L)), vnam = "EduField")
```
\emini


- Observed factor levels: \"Human Resources\", \"Life Sciences\", \"Marketing\", \"Medical\", \"Other\", \"Technical Degree\". 



\fullline

### EmpCount

* The variable only takes one (non-missing) value: \"1\". The variable contains 0 \% missing observations.



\fullline

### EmpId

\bminione

-------------------------------------------
Feature                              Result
------------------------- -----------------
Variable type                       numeric

Number of missing obs.              0 (0 %)

Number of unique values                1470

Median                               1020.5

1st and 3rd quartiles       491.25; 1555.75

Min. and max.                       1; 2068
-------------------------------------------


\emini
\bminitwo
```{r 'Var-10-EmpId', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(0, 200, 400, 600, 800, 
1000, 1200, 1400, 1600, 1800, 2000), xmax = c(200, 400, 600, 
800, 1000, 1200, 1400, 1600, 1800, 2000, 2200), ymin = c(0, 0, 
0, 0, 0, 0, 0, 0, 0, 0, 0), ymax = c(149L, 144L, 155L, 129L, 
140L, 141L, 136L, 140L, 151L, 137L, 48L)), class = "data.frame", row.names = c(NA, 
-11L)), vnam = "EmpId")
```
\emini




\fullline

### EnvironSat

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values           4

Median                            3

1st and 3rd quartiles          2; 4

Min. and max.                  1; 4
-----------------------------------


\emini
\bminitwo
```{r 'Var-11-EnvironSat', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(1, 1.2, 1.4, 1.6, 1.8, 
2, 2.2, 2.4, 2.6, 2.8, 3, 3.2, 3.4, 3.6, 3.8), xmax = c(1.2, 
1.4, 1.6, 1.8, 2, 2.2, 2.4, 2.6, 2.8, 3, 3.2, 3.4, 3.6, 3.8, 
4), ymin = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0), ymax = c(284L, 
0L, 0L, 0L, 287L, 0L, 0L, 0L, 0L, 453L, 0L, 0L, 0L, 0L, 446L)), class = "data.frame", row.names = c(NA, 
-15L)), vnam = "EnvironSat")
```
\emini




\fullline

### Gender

\bminione

-------------------------------------
Feature                        Result
------------------------- -----------
Variable type               character

Number of missing obs.        0 (0 %)

Number of unique values             2

Mode                           "Male"
-------------------------------------


\emini
\bminitwo
```{r 'Var-12-Gender', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggBarplot(data = structure(list(x = structure(1:2, .Label = c("Female", 
"Male"), class = "factor"), y = c(588L, 882L)), class = "data.frame", row.names = c(NA, 
-2L)), vnam = "Gender")
```
\emini


- Observed factor levels: \"Female\", \"Male\". 



\fullline

### HourlyRate

\bminione

-------------------------------------
Feature                        Result
------------------------- -----------
Variable type                 numeric

Number of missing obs.        0 (0 %)

Number of unique values            71

Median                             66

1st and 3rd quartiles       48; 83.75

Min. and max.                 30; 100
-------------------------------------


\emini
\bminitwo
```{r 'Var-13-HourlyRate', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(30L, 35L, 40L, 45L, 
50L, 55L, 60L, 65L, 70L, 75L, 80L, 85L, 90L, 95L), xmax = c(35L, 
40L, 45L, 50L, 55L, 60L, 65L, 70L, 75L, 80L, 85L, 90L, 95L, 100L
), ymin = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0), ymax = c(107L, 
84L, 115L, 103L, 104L, 114L, 97L, 93L, 107L, 111L, 113L, 103L, 
104L, 115L)), class = "data.frame", row.names = c(NA, -14L)), 
    vnam = "HourlyRate")
```
\emini




\fullline

### JobInvolvemt

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values           4

Median                            3

1st and 3rd quartiles          2; 3

Min. and max.                  1; 4
-----------------------------------


\emini
\bminitwo
```{r 'Var-14-JobInvolvemt', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(1, 1.2, 1.4, 1.6, 1.8, 
2, 2.2, 2.4, 2.6, 2.8, 3, 3.2, 3.4, 3.6, 3.8), xmax = c(1.2, 
1.4, 1.6, 1.8, 2, 2.2, 2.4, 2.6, 2.8, 3, 3.2, 3.4, 3.6, 3.8, 
4), ymin = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0), ymax = c(83L, 
0L, 0L, 0L, 375L, 0L, 0L, 0L, 0L, 868L, 0L, 0L, 0L, 0L, 144L)), class = "data.frame", row.names = c(NA, 
-15L)), vnam = "JobInvolvemt")
```
\emini




\fullline

### JobLevel

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values           5

Median                            2

1st and 3rd quartiles          1; 3

Min. and max.                  1; 5
-----------------------------------


\emini
\bminitwo
```{r 'Var-15-JobLevel', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(1, 1.5, 2, 2.5, 3, 3.5, 
4, 4.5), xmax = c(1.5, 2, 2.5, 3, 3.5, 4, 4.5, 5), ymin = c(0, 
0, 0, 0, 0, 0, 0, 0), ymax = c(543L, 534L, 0L, 218L, 0L, 106L, 
0L, 69L)), class = "data.frame", row.names = c(NA, -8L)), vnam = "JobLevel")
```
\emini




\fullline

### JobRole

\bminione

---------------------------------------------
Feature                                Result
------------------------- -------------------
Variable type                       character

Number of missing obs.                0 (0 %)

Number of unique values                     9

Mode                        "Sales Executive"
---------------------------------------------


\emini
\bminitwo
```{r 'Var-16-JobRole', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggBarplot(data = structure(list(x = structure(1:9, .Label = c("Healthcare Representative", 
"Human Resources", "Laboratory Technician", "Manager", "Manufacturing Director", 
"Research Director", "Research Scientist", "Sales Executive", 
"Sales Representative"), class = "factor"), y = c(131L, 52L, 
259L, 102L, 145L, 80L, 292L, 326L, 83L)), class = "data.frame", row.names = c(NA, 
-9L)), vnam = "JobRole")
```
\emini


- Observed factor levels: \"Healthcare Representative\", \"Human Resources\", \"Laboratory Technician\", \"Manager\", \"Manufacturing Director\", \"Research Director\", \"Research Scientist\", \"Sales Executive\", \"Sales Representative\". 



\fullline

### JobSat

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values           4

Median                            3

1st and 3rd quartiles          2; 4

Min. and max.                  1; 4
-----------------------------------


\emini
\bminitwo
```{r 'Var-17-JobSat', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(1, 1.2, 1.4, 1.6, 1.8, 
2, 2.2, 2.4, 2.6, 2.8, 3, 3.2, 3.4, 3.6, 3.8), xmax = c(1.2, 
1.4, 1.6, 1.8, 2, 2.2, 2.4, 2.6, 2.8, 3, 3.2, 3.4, 3.6, 3.8, 
4), ymin = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0), ymax = c(289L, 
0L, 0L, 0L, 280L, 0L, 0L, 0L, 0L, 442L, 0L, 0L, 0L, 0L, 459L)), class = "data.frame", row.names = c(NA, 
-15L)), vnam = "JobSat")
```
\emini




\fullline

### MaritalStat

\bminione

-------------------------------------
Feature                        Result
------------------------- -----------
Variable type               character

Number of missing obs.        0 (0 %)

Number of unique values             3

Mode                        "Married"
-------------------------------------


\emini
\bminitwo
```{r 'Var-18-MaritalStat', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggBarplot(data = structure(list(x = structure(1:3, .Label = c("Divorced", 
"Married", "Single"), class = "factor"), y = c(327L, 673L, 470L
)), class = "data.frame", row.names = c(NA, -3L)), vnam = "MaritalStat")
```
\emini


- Observed factor levels: \"Divorced\", \"Married\", \"Single\". 



\fullline

### MnthInc

\bminione

---------------------------------------
Feature                          Result
------------------------- -------------
Variable type                   numeric

Number of missing obs.          0 (0 %)

Number of unique values            1349

Median                             4919

1st and 3rd quartiles        2911; 8379

Min. and max.               1009; 19999
---------------------------------------


\emini
\bminitwo
```{r 'Var-19-MnthInc', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(0, 2000, 4000, 6000, 
8000, 10000, 12000, 14000, 16000, 18000), xmax = c(2000, 4000, 
6000, 8000, 10000, 12000, 14000, 16000, 18000, 20000), ymin = c(0, 
0, 0, 0, 0, 0, 0, 0, 0, 0), ymax = c(33L, 510L, 371L, 175L, 100L, 
86L, 53L, 16L, 57L, 69L)), class = "data.frame", row.names = c(NA, 
-10L)), vnam = "MnthInc")
```
\emini




\fullline

### MnthRate

\bminione

-----------------------------------------
Feature                            Result
------------------------- ---------------
Variable type                     numeric

Number of missing obs.            0 (0 %)

Number of unique values              1427

Median                            14235.5

1st and 3rd quartiles       8047; 20461.5

Min. and max.                 2094; 26999
-----------------------------------------


\emini
\bminitwo
```{r 'Var-20-MnthRate', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(2000L, 4000L, 6000L, 
8000L, 10000L, 12000L, 14000L, 16000L, 18000L, 20000L, 22000L, 
24000L, 26000L), xmax = c(4000L, 6000L, 8000L, 10000L, 12000L, 
14000L, 16000L, 18000L, 20000L, 22000L, 24000L, 26000L, 28000L
), ymin = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0), ymax = c(110L, 
130L, 123L, 124L, 118L, 116L, 114L, 120L, 117L, 119L, 131L, 99L, 
49L)), class = "data.frame", row.names = c(NA, -13L)), vnam = "MnthRate")
```
\emini




\fullline

### NumCoWorked

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values          10

Median                            2

1st and 3rd quartiles          1; 4

Min. and max.                  0; 9
-----------------------------------


\emini
\bminitwo
```{r 'Var-21-NumCoWorked', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(0, 1, 2, 3, 4, 5, 6, 
7, 8), xmax = c(1, 2, 3, 4, 5, 6, 7, 8, 9), ymin = c(0, 0, 0, 
0, 0, 0, 0, 0, 0), ymax = c(718L, 146L, 159L, 139L, 63L, 70L, 
74L, 49L, 52L)), class = "data.frame", row.names = c(NA, -9L)), 
    vnam = "NumCoWorked")
```
\emini




\fullline

### Over18

* The variable only takes one (non-missing) value: \"Y\". The variable contains 0 \% missing observations.



\fullline

### Overtime

\bminione

-------------------------------------
Feature                        Result
------------------------- -----------
Variable type               character

Number of missing obs.        0 (0 %)

Number of unique values             2

Mode                             "No"
-------------------------------------


\emini
\bminitwo
```{r 'Var-23-Overtime', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggBarplot(data = structure(list(x = structure(1:2, .Label = c("No", 
"Yes"), class = "factor"), y = c(1054L, 416L)), class = "data.frame", row.names = c(NA, 
-2L)), vnam = "Overtime")
```
\emini


- Observed factor levels: \"No\", \"Yes\". 



\fullline

### PcntSalInc

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values          15

Median                           14

1st and 3rd quartiles        12; 18

Min. and max.                11; 25
-----------------------------------


\emini
\bminitwo
```{r 'Var-24-PcntSalInc', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = 11:24, xmax = 12:25, ymin = c(0, 
0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0), ymax = c(408L, 209L, 
201L, 101L, 78L, 82L, 89L, 76L, 55L, 48L, 56L, 28L, 21L, 18L)), class = "data.frame", row.names = c(NA, 
-14L)), vnam = "PcntSalInc")
```
\emini




\fullline

### PerfRating

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values           2

Median                            3

1st and 3rd quartiles          3; 3

Min. and max.                  3; 4
-----------------------------------


\emini
\bminitwo
```{r 'Var-25-PerfRating', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(3, 3.1, 3.2, 3.3, 3.4, 
3.5, 3.6, 3.7, 3.8, 3.9), xmax = c(3.1, 3.2, 3.3, 3.4, 3.5, 3.6, 
3.7, 3.8, 3.9, 4), ymin = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0), ymax = c(1244L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 226L)), class = "data.frame", row.names = c(NA, 
-10L)), vnam = "PerfRating")
```
\emini




\fullline

### RelationSat

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values           4

Median                            3

1st and 3rd quartiles          2; 4

Min. and max.                  1; 4
-----------------------------------


\emini
\bminitwo
```{r 'Var-26-RelationSat', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(1, 1.2, 1.4, 1.6, 1.8, 
2, 2.2, 2.4, 2.6, 2.8, 3, 3.2, 3.4, 3.6, 3.8), xmax = c(1.2, 
1.4, 1.6, 1.8, 2, 2.2, 2.4, 2.6, 2.8, 3, 3.2, 3.4, 3.6, 3.8, 
4), ymin = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0), ymax = c(276L, 
0L, 0L, 0L, 303L, 0L, 0L, 0L, 0L, 459L, 0L, 0L, 0L, 0L, 432L)), class = "data.frame", row.names = c(NA, 
-15L)), vnam = "RelationSat")
```
\emini




\fullline

### StandHours

* The variable only takes one (non-missing) value: \"80\". The variable contains 0 \% missing observations.



\fullline

### StckOptLev

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values           4

Median                            1

1st and 3rd quartiles          0; 1

Min. and max.                  0; 3
-----------------------------------


\emini
\bminitwo
```{r 'Var-28-StckOptLev', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(0, 0.2, 0.4, 0.6, 0.8, 
1, 1.2, 1.4, 1.6, 1.8, 2, 2.2, 2.4, 2.6, 2.8), xmax = c(0.2, 
0.4, 0.6, 0.8, 1, 1.2, 1.4, 1.6, 1.8, 2, 2.2, 2.4, 2.6, 2.8, 
3), ymin = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0), ymax = c(631L, 
0L, 0L, 0L, 596L, 0L, 0L, 0L, 0L, 158L, 0L, 0L, 0L, 0L, 85L)), class = "data.frame", row.names = c(NA, 
-15L)), vnam = "StckOptLev")
```
\emini




\fullline

### TotalWrkYrs

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values          40

Median                           10

1st and 3rd quartiles         6; 15

Min. and max.                 0; 40
-----------------------------------


\emini
\bminitwo
```{r 'Var-29-TotalWrkYrs', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(0, 5, 10, 15, 20, 25, 
30, 35), xmax = c(5, 10, 15, 20, 25, 30, 35, 40), ymin = c(0, 
0, 0, 0, 0, 0, 0, 0), ymax = c(316L, 607L, 191L, 149L, 109L, 
52L, 33L, 13L)), class = "data.frame", row.names = c(NA, -8L)), 
    vnam = "TotalWrkYrs")
```
\emini




\fullline

### TrainPrevYr

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values           7

Median                            3

1st and 3rd quartiles          2; 3

Min. and max.                  0; 6
-----------------------------------


\emini
\bminitwo
```{r 'Var-30-TrainPrevYr', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(0, 0.5, 1, 1.5, 2, 2.5, 
3, 3.5, 4, 4.5, 5, 5.5), xmax = c(0.5, 1, 1.5, 2, 2.5, 3, 3.5, 
4, 4.5, 5, 5.5, 6), ymin = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 
0), ymax = c(54L, 71L, 0L, 547L, 0L, 491L, 0L, 123L, 0L, 119L, 
0L, 65L)), class = "data.frame", row.names = c(NA, -12L)), vnam = "TrainPrevYr")
```
\emini




\fullline

### WorkLifeBal

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values           4

Median                            3

1st and 3rd quartiles          2; 3

Min. and max.                  1; 4
-----------------------------------


\emini
\bminitwo
```{r 'Var-31-WorkLifeBal', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(1, 1.2, 1.4, 1.6, 1.8, 
2, 2.2, 2.4, 2.6, 2.8, 3, 3.2, 3.4, 3.6, 3.8), xmax = c(1.2, 
1.4, 1.6, 1.8, 2, 2.2, 2.4, 2.6, 2.8, 3, 3.2, 3.4, 3.6, 3.8, 
4), ymin = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0), ymax = c(80L, 
0L, 0L, 0L, 344L, 0L, 0L, 0L, 0L, 893L, 0L, 0L, 0L, 0L, 153L)), class = "data.frame", row.names = c(NA, 
-15L)), vnam = "WorkLifeBal")
```
\emini




\fullline

### YrsAtComp

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values          37

Median                            5

1st and 3rd quartiles          3; 9

Min. and max.                 0; 40
-----------------------------------


\emini
\bminitwo
```{r 'Var-32-YrsAtComp', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(0, 5, 10, 15, 20, 25, 
30, 35), xmax = c(5, 10, 15, 20, 25, 30, 35, 40), ymin = c(0, 
0, 0, 0, 0, 0, 0, 0), ymax = c(776L, 448L, 108L, 72L, 41L, 9L, 
12L, 4L)), class = "data.frame", row.names = c(NA, -8L)), vnam = "YrsAtComp")
```
\emini




\fullline

### YrsCurrRole

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values          19

Median                            3

1st and 3rd quartiles          2; 7

Min. and max.                 0; 18
-----------------------------------


\emini
\bminitwo
```{r 'Var-33-YrsCurrRole', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(0, 2, 4, 6, 8, 10, 12, 
14, 16), xmax = c(2, 4, 6, 8, 10, 12, 14, 16, 18), ymin = c(0, 
0, 0, 0, 0, 0, 0, 0, 0), ymax = c(673L, 239L, 73L, 311L, 96L, 
32L, 25L, 15L, 6L)), class = "data.frame", row.names = c(NA, 
-9L)), vnam = "YrsCurrRole")
```
\emini




\fullline

### YrsSncPromo

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values          16

Median                            1

1st and 3rd quartiles          0; 3

Min. and max.                 0; 15
-----------------------------------


\emini
\bminitwo
```{r 'Var-34-YrsSncPromo', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(0, 1, 2, 3, 4, 5, 6, 
7, 8, 9, 10, 11, 12, 13, 14), xmax = c(1, 2, 3, 4, 5, 6, 7, 8, 
9, 10, 11, 12, 13, 14, 15), ymin = c(0, 0, 0, 0, 0, 0, 0, 0, 
0, 0, 0, 0, 0, 0, 0), ymax = c(938L, 159L, 52L, 61L, 45L, 32L, 
76L, 18L, 17L, 6L, 24L, 10L, 10L, 9L, 13L)), class = "data.frame", row.names = c(NA, 
-15L)), vnam = "YrsSncPromo")
```
\emini




\fullline

### YrsWtCurMgr

\bminione

-----------------------------------
Feature                      Result
------------------------- ---------
Variable type               numeric

Number of missing obs.      0 (0 %)

Number of unique values          18

Median                            3

1st and 3rd quartiles          2; 7

Min. and max.                 0; 17
-----------------------------------


\emini
\bminitwo
```{r 'Var-35-YrsWtCurMgr', echo=FALSE, fig.width=4, fig.height=3, message=FALSE, warning=FALSE}
ggAggHist(data = structure(list(xmin = c(0, 2, 4, 6, 8, 10, 12, 
14, 16), xmax = c(2, 4, 6, 8, 10, 12, 14, 16, 18), ymin = c(0, 
0, 0, 0, 0, 0, 0, 0, 0), ymax = c(683L, 240L, 60L, 323L, 91L, 
40L, 19L, 7L, 7L)), class = "data.frame", row.names = c(NA, -9L
)), vnam = "YrsWtCurMgr")
```
\emini




\fullline

# **Analysis Objects**
Two versions of the original employee data frame are created. One converts all character variables and select rating-type numeric variables to factors. The other goes a step further and coerces all variables to numeric. Both are used in different ways during the analysis.

### employee_factor

---------------------------------------------------
Variable             Class       # unique  Missing  
                                           values                        
-------------------- --------- ---------- --------- 
**[Age]**            numeric           43  0.00 %                

**[Attrition]**      factor             2  0.00 %                

**[BusTravel]**      factor             3  0.00 %                

**[DailyRate]**      numeric          886  0.00 %                

**[Department]**     factor             3  0.00 %                

**[DistFromHme]**    numeric           29  0.00 %                

**[Education]**      factor             5  0.00 %                

**[EduField]**       factor             6  0.00 %                

**[EmpCount]**       numeric            1  0.00 %                

**[EmpId]**          numeric         1470  0.00 %                

**[EnvironSat]**     factor             4  0.00 %                

**[Gender]**         factor             2  0.00 %                

**[HourlyRate]**     numeric           71  0.00 %                

**[JobInvolvemt]**   factor             4  0.00 %                

**[JobLevel]**       factor             5  0.00 %                

**[JobRole]**        factor             9  0.00 %                

**[JobSat]**         factor             4  0.00 %                

**[MaritalStat]**    factor             3  0.00 %                

**[MnthInc]**        numeric         1349  0.00 %                

**[MnthRate]**       numeric         1427  0.00 %                

**[NumCoWorked]**    numeric           10  0.00 %                

**[Over18]**         factor             1  0.00 %                

**[Overtime]**       factor             2  0.00 %                

**[PcntSalInc]**     numeric           15  0.00 %                

**[PerfRating]**     factor             2  0.00 %                

**[RelationSat]**    factor             4  0.00 %                

**[StandHours]**     factor             1  0.00 %                

**[StckOptLev]**     factor             4  0.00 %                

**[TotalWrkYrs]**    numeric           40  0.00 %                

**[TrainPrevYr]**    numeric            7  0.00 %                

**[WorkLifeBal]**    factor             4  0.00 %                

**[YrsAtComp]**      numeric           37  0.00 %                

**[YrsCurrRole]**    numeric           19  0.00 %                

**[YrsSncPromo]**    numeric           16  0.00 %                

**[YrsWtCurMgr]**    numeric           18  0.00 %                
-------------------------------------------------------------------------


### employee_numeric


---------------------------------------------------
Variable             Class       # unique  Missing  
                                           values                        
-------------------- --------- ---------- --------- 
**[Age]**            numeric           43  0.00 %                

**[Attrition]**      numeric             2  0.00 %                

**[BusTravel]**      numeric             3  0.00 %                

**[DailyRate]**      numeric          886  0.00 %                

**[Department]**     numeric             3  0.00 %                

**[DistFromHme]**    numeric           29  0.00 %                

**[Education]**      numeric             5  0.00 %                

**[EduField]**       numeric             6  0.00 %                

**[EmpCount]**       numeric            1  0.00 %                

**[EmpId]**          numeric         1470  0.00 %                

**[EnvironSat]**     numeric             4  0.00 %                

**[Gender]**         numeric             2  0.00 %                

**[HourlyRate]**     numeric           71  0.00 %                

**[JobInvolvemt]**   numeric             4  0.00 %                

**[JobLevel]**       numeric             5  0.00 %                

**[JobRole]**        numeric             9  0.00 %                

**[JobSat]**         numeric             4  0.00 %                

**[MaritalStat]**    numeric             3  0.00 %                

**[MnthInc]**        numeric         1349  0.00 %                

**[MnthRate]**       numeric         1427  0.00 %                

**[NumCoWorked]**    numeric           10  0.00 %                

**[Over18]**         numeric             1  0.00 %                

**[Overtime]**       numeric             2  0.00 %                

**[PcntSalInc]**     numeric           15  0.00 %                

**[PerfRating]**     numeric             2  0.00 %                

**[RelationSat]**    numeric             4  0.00 %                

**[StandHours]**     numeric             1  0.00 %                

**[StckOptLev]**     numeric             4  0.00 %                

**[TotalWrkYrs]**    numeric           40  0.00 %                

**[TrainPrevYr]**    numeric            7  0.00 %                

**[WorkLifeBal]**    numeric             4  0.00 %                

**[YrsAtComp]**      numeric           37  0.00 %                

**[YrsCurrRole]**    numeric           19  0.00 %                

**[YrsSncPromo]**    numeric           16  0.00 %                

**[YrsWtCurMgr]**    numeric           18  0.00 %                
-------------------------------------------------------------------------

# Non-Visualization Variables

 

\fullline

-----------------------------------------------------------------------------------
Feature                      Result
-------------------------- --------------------------------------------------------
**descriptiveStats1**       Age, MnthInc, PcntSalInc, YrsAtComp  

**descriptiveStats2**       YrsCurrRole, YrsSncPromo, YrsWtCurMgr

**EducationData**           Below College, College, Bachelor, Master, Doctor

**GenderData**              Male, Female 

**JobRoleData**             Healthcare Representative, Human Resources, Laboratory Technician, Manager                                        Manufacturing Director, Research Director, Research Scientist, 
                            Sales Executive, Sales Representative

**ManagerRoleData**         Manager, Manufacturing Director, Research Director

**lm1**                     linear model for MnthInc and Age

**lm2**                     linear model for log of MnthInc and Age

**employee_log**            copy of employee_factor with log of the MnthInc variable

**train_rows**              data partition of employee_factor

**train_subset**            Subset of employee data for training model

**test_subset**             Subset of employee data for test model 

**null_model**              Model with no predictors

**full_model**              Mode with full predictors

**glmStep_model**           Stepwise regression model 






-----------------------------------------------------------------------------------


# Visualization Variables

 

\fullline

-----------------------------------------------------------------------------------
Feature                      Result
-------------------------- --------------------------------------------------------
**ggAge**             Histogram of Age

**ggIncome**          Histogram of Income 

**ggpairs**           Visualize the relationship between rates and income 
                      HourlyRate, DailyRate, MnthRate, MnthInc 

**employee_money**    upper and lower limits for the rates HourlyRate, DailyRate, MnthRate, MnthInc



**JobRoleData**             Healthcare Representative, Human Resources, Laboratory Technician, Manager                                        Manufacturing Director, Research Director, Research Scientist, 
                            Sales Executive, Sales Representative

**satisfaction**            "JobSat", "WorkLifeBal", "RelationSat"
**ManagerRoleData**         Manager, Manufacturing Director, Research Director

**jobrole_age_a**           Throwaway graph to obtain common job role

**jobrole_monthincome**     Job role and monthly income boxplots

**jobrole_salaryincrease**  Job role and percent salary increase

**jobrole_distancehome**    Job role and distance from home boxplots

**jobrole_age**             Job role and Age boxplots

**jobrole_totalworkyears**  Job role and working years boxplots

**jobrole_currentyears**    Job role and years in current role boxplot

**jobrole_yearsatcomp**     Job role and years at company boxplot

**jobrole_currentmamnager** Job role and years with current manager boxplot

**jobrole_numcompanies**    Job role and numb of companies worked boxplot

**jobrole_totalworkyears**  Job role and total work years boxplot

**jobrole_currentyears**    Job role and years in current role boxplot

**jobrole_numcompanies***   Job role and number of companies worked

**job_attrition**           Job role and attrition bar graphs

**job_travel**              Job role and travel bar graphs

**job_dept**                Job role and dept bar graphs

**job_edufield**            Job role and education bar graphs

**job_marital**             Job role and marital status bar graphs

**job_overttime**           Job role and overtime status bar graphs

**job_att_percent_bar**     Percentage within job role for attrition and overttime

**att_age_a**               Throwaway graph to obtain common Att

**att_age**                 Attrition and Age density plot

**att_distance**            Attrition and distance from home density plot

**att_monthlyincome**       Attrition and monthly income density plot

**att_salaryincrease**      Attrition and percent salary increase density plot

**att_yearscomp**           Attrition and Years in current role density plot 

**att_yearsrole**           Attrition in current role  density plot

**att_yearspromo**          Attrition and years since last promotion density plot

**att_yearsmanager**        Attrition and years with current manager  desity plot

**att_travel_count**        Attrition and business travel bar plot

**att_dept_count**          Attrition and department bar plot

**att_edufied_count**       Attrition and Education field bar plot

**att_gender_count**        Attrition and Gender field bar plot

**att_marital_count**       Attrition and Marital field bar plot 

**att_overtime_count**      Attrition and Overtime field bar plot 

**knn_model**               Train model for knn

**prediction_knn**          Predict for knn model 

**cart_model**              Classification and Regression tree model on training

**employee_rf**             Random forest model for training data


-----------------------------------------------------------------------------------